---
title: "Working with CAMS solar data"
author: "Lukas Lundström"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with CAMS solar data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r echo=FALSE}
knitr::opts_chunk$set(
  comment = "#>", 
  collapse = TRUE,
  warning = FALSE
)
```

## CAMS radiation service
The CAMS radiation service provides time series of global, direct, and diffuse Irradiations on horizontal surface, and direct Irradiation on normal plane for the actual weather conditions as well as for clear-sky conditions. The data can be accessed manually on the CAMS radiation service [site](http://www.soda-pro.com/web-services/radiation/cams-radiation-service). The service is part of the [Copernicus Atmosphere Monitoring Service](http://atmosphere.copernicus.eu/) (CAMS).

* The geographical coverage is the field-of-view of the Meteosat satellite, roughly speaking Europe, Africa, Atlantic Ocean, Middle East (-66° to 66° in both latitudes and longitudes).
* The time coverage of data is from 2004-02-01 up to 2 days ago. Data are available with a time step ranging from 15 min to 1 month. The service allows time steps of 1 min as well, these will be interpolated values from the 15 min time steps (accounting for changes in clear-sky conditions). 
* The horizontal resolution is the original pixel of the Meteosat images, 3-5 km.  The CAMS radiation service has currently (`r Sys.Date()`) limited the amount of requests to 15 per day. This limit may evolve. 

## Authentication

To access the CAMS radiation service you need to register at their [site](http://www.soda-pro.com/web-services/radiation/cams-radiation-service). The email you use at the registration step will be used for authentication. Set the environment variable `CAMS_USERNAME` to your registered email to follow the examples below: `Sys.setenv(CAMS_USERNAME="your@email.com")`

## Retrieve hourly solar data

`cams_get_radiation()` and `cams_get_radiation()` are convenience wrappers that retrieves CAMS solar data straight into a R data frame. The example bellow retrieves hourly radiation data for the location 60° latitude and 15° longitude for the period 2016-01-01 to 2016-01-15.
```{r}  
library(camsRad)

df <- cams_get_radition(
  Sys.getenv("CAMS_USERNAME"), # Authentication
  lat=60, lng=15, # Latitude/longitude coordinates 
  date_begin="2016-01-01", date_end="2016-01-15", # Date range
  time_step="PT01H") # Use hourly time step

```
As seen above the `cams_get_radiation()` prints additional information about the data, these can be suppressed by wrapping the call with `suppressMessages()`. 

```{r}
print(df)

```
The first column holds the timestamp information. It follows the convention of giving solar radiation as the sum during the previous hour. E.g. the timestamp of 14:00 shows the solar radiation during 13:00-14:00. 

## More advanced retrievals
To use other data formats and to save data to the disk we need to use the `cams_api()`. The example bellow writes daily solar radiation in netCDF format to the disk. You need to have the `ncdf4` package installed.

```{r, warning = FALSE, fig.width=7}
library(ncdf4)

filename <- paste0(tempfile(), ".nc")

r <- cams_api(
  Sys.getenv("CAMS_USERNAME"), 60, 15, "2016-06-01", "2016-07-1", 
  format="application/x-netcdf",
  time_step = "P01D",
  filename=filename)

# Access the on disk stored netCDF file
nc <- nc_open(filename)  

# List names of available variables
names(nc$var)

# Create data.frame with datetime and global horizontal irradiation and plot it
df <- data.frame(timestamp = as.POSIXct(nc$dim$time$vals, "UTC", origin="1970-01-01"),
                 GHI = ncvar_get(nc, "GHI"))
df$timestamp <- df$timestamp-3600*24 # shift timestamp 24 hours backwards

nc_close(nc) 

plot(df, type="b", ylab="GHI, Wh/m2,day", xlab="2016")


```
Note that the *timestamp* follows the convention of giving solar radiation as the sum during the previous time step. This is usually correct when working with hourly data. But when working with daily (or monthly) data it is more relevant to have the *timestamp* at the starting point of summation. The `df$timestamp-3600*24`part achieves this for daily data.

To get the data in *csv* or *json* format instead of *netCDF*, just change the *format* parameter to "application/csv" or "application/json" (and the filename extension to *.csv* or *.json* respectively).
